---
output: html_document
params:
  i: 1
---


```{r include = FALSE}
library(tidyverse)
library(leaflet)
load("assets/cases_db.Rda")
x <- dat$id[params$i]
cs <- dat |> filter(id== x)
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```

# `r cs$case_study_name`

**Main contributors**: `r cs |> pull(main_contributors)`

**Other contributors**: `r cs |> pull(other_contributors)`

Last update: `r cs$date`



```{r map, out.width="100%", out.height="300px", results = 'markup'}
m <- leaflet(cs ) |> 
    addTiles("https://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png") |> 
    addMarkers(cs$long, cs$lat) |> 
    setView(cs$long, cs$lat, zoom = 4)
m
```


### Summary

```{r summary, results='asis'}
cs |> pull(summary) |> cat()
```

**Type of regime shift**: `r dat |> filter(id == x) |> pull(type) `

**Ecosystem type**:

```{r results='asis'}
cs|> pull(ecosystem_type) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Land uses**:

```{r results='asis'}
cs |> pull(land_uses) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Spatial scale**:

```{r results='asis'}
cs |> pull(spatial_scale) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Continent or Ocean**:

```{r results='asis'}
cs |> pull(location_continent_or_ocean) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Region**: `r cs|> pull(location_region) `

**Countries**: `r cs |> pull(location_countries) `


{.tabset}
--------------------------------------------------------------------------------

### Categorical attributes

::::{style="display: flex;"}


:::{style="width: 50%;"}

#### Impacts

**Ecosystem type**:

```{r results='asis'}
cs |> pull(ecosystem_type) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Key ecosystem processes**:

```{r results='asis'}
cs |> pull(impacts_key_ecosystem_processes) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Biodiversity**:

```{r results='asis'}
cs |> pull(impacts_biodiversity) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Provisioning services**:

```{r results='asis'}
cs |> pull(impacts_provisioning_services) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Regulating services**:

```{r results='asis'}
cs |> pull(impacts_regulating_services) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Cultural services**:

```{r results='asis'}
cs |> pull(impacts_cultural_services) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Human well-being**:

```{r results='asis'}
cs |> pull(impacts_on_human_well_being) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```


:::

:::{style="width: 50%;"}

#### Drivers

**Key drivers**:

```{r results='asis'}
cs |> pull(drivers_key_direct_drivers) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Land use**:

```{r results='asis'}
cs |> pull(land_uses) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```


#### Key attributes

**Spatial scale**:

```{r results='asis'}
cs |> pull(key_attributes_spatial_scale_of_rs) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Time scale**:

```{r results='asis'}
cs |> pull(key_attributes_time_scale_of_rs) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Reversibility**:

```{r results='asis'}
cs |> pull(key_attributes_reversibility) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Evidence**:

```{r results='asis'}
cs |> pull(key_attributes_evidence) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Confidence: existence of the regime shift**

```{r results='asis'}
cs |> pull(key_attributes_confidence_existence_of_rs) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Confidence: mechanisms underlying the regime shift**

```{r results='asis'}
cs |> pull(key_attributes_confidence_mechanism_underlying_rs) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

:::

::::

### Detail info and key references

#### Alternative regimes

```{r results='asis'}
cs |> pull(alternate_regimes) |> str_remove_all(pattern = "\\<br \\/\\>") |>  cat()
```

#### Drivers and causes of the regime shift

```{r results='asis'}
cs |> pull(drivers_and_causes_of_the_regime_shift) |> 
    str_remove_all(pattern = "\\<p\\>\\&nbsp\\;\\<\\/p\\>") |> cat()
```

#### Impacts on ecosystem services and human well-being

```{r results='asis'}
cs |> pull(impacts_on_ecosystem_services_and_human_well_being) |> cat()
```

#### Management options

```{r results='asis'}
cs |> pull(management_options) |> cat()
```


#### Key references

<small>

```{r results='asis'}
cs |> pull(references) |> 
    str_split(pattern = "\\(\\#\\d{1,}\\) ") |> 
    map(~str_c("- ", ., "\n")) |> 
    unlist() |> 
    str_remove(pattern = "- \\\n") |> 
    cat()
```

</small>


