---
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true

params:
  i: 7
---

```{r include = FALSE}
library(tidyverse)
library(leaflet)
load('assets/cases_db.Rda')
cases <- dat
dat <- read_csv("assets/generic_types_RSDB_new.csv") |> 
    janitor::clean_names()
x <- dat$id[params$i]
rs <- dat |> filter(id== x)
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```

# `r rs$regime_shift_name`

**Main contributors**: `r rs |> pull(main_contributors)`

**Other contributors**: `r rs |> pull(other_contributors)`

Last update: `r rs$date`




```{r map, out.width="100%", out.height="300px", results = 'markup'}

cs <- cases |> 
    filter(type == rs$regime_shift_name) 

m <- leaflet(cs) |> 
    addTiles("https://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png") |> 
    addCircleMarkers(lng = ~long, lat=~lat, color = "orange", radius = 2, popup = ~popups) |> 
    setView(5,10, zoom = 2) 
m
```

### Summary

```{r summary, results='asis'}
rs |> pull(summary) |> cat()
```

### Categorical attributes

::: {style="display: flex;"}
::: {style="width: 50%;"}
#### Impacts

**Ecosystem type**:

```{r results='asis'}
rs |> pull(impacts_ecosystem_type) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Key ecosystem processes**:

```{r results='asis'}
rs |> pull(impacts_key_ecosystem_processes) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Biodiversity**:

```{r results='asis'}
rs |> pull(impacts_biodiversity) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Provisioning services**:

```{r results='asis'}
rs |> pull(impacts_provisioning_services) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Regulating services**:

```{r results='asis'}
rs |> pull(impacts_regulating_services) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Cultural services**:

```{r results='asis'}
rs |> pull(impacts_cultural_services) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Human well-being**:

```{r results='asis'}
rs |> pull(impacts_human_well_being) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Links to other regime shifts**

```{r results='asis'}
rs |> pull(links_to_other_regime_shifts) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```
:::

::: {style="width: 50%;"}
#### Drivers

**Key drivers**:

```{r results='asis'}
rs |> pull(drivers_key_direct_drivers) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Land use**:

```{r results='asis'}
rs |> pull(drivers_land_use) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

#### Key attributes

**Spatial scale**:

```{r results='asis'}
rs |> pull(key_attributes_typical_spatial_scale) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Time scale**:

```{r results='asis'}
rs |> pull(key_attributes_typical_time_scale) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Reversibility**:

```{r results='asis'}
rs |> pull(key_attributes_reversibility) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Evidence**:

```{r results='asis'}
rs |> pull(key_attributes_evidence) |> str_split(pattern = ", ") |> 
    map(~str_c("- ", ., "\n")) |> unlist() |> cat()
```

**Confidence: existence of the regime shift**

```{r results='asis'}
rs |> pull(key_attributes_confidence_existence_of_rs) %>% str_c("- ", . , "\n") |> cat()
```

**Confidence: mechanisms underlying the regime shift**

```{r results='asis'}
rs |> pull(key_attributes_confidence_mechanism_underlying_rs)  %>% str_c("- ", . , "\n") |> cat()
```
:::
:::

<!-- ##  {.tabset .tabset-fades .tabset-pills} -->

### Detail info

#### Alternative regimes

```{r results='asis'}
rs |> pull(alternate_regimes) |> str_remove_all(pattern = "\\<br \\/\\>") |>  cat()
```

#### Drivers and causes of the regime shift

```{r results='asis'}
rs |> pull(drivers_and_causes_of_the_regime_shift) |> 
    str_remove_all(pattern = "\\<p\\>\\&nbsp\\;\\<\\/p\\>") |> cat()
```

#### Impacts on ecosystem services and human well-being

```{r results='asis'}
rs |> pull(impacts_on_ecosystem_services_and_human_well_being) |> cat()
```

#### Management options

```{r results='asis'}
rs |> pull(management_options) |> cat()
```


### Regime shift Analysis

```{r cld, error = FALSE, message = FALSE, out.width="80%"}
source("tools.R")
load("assets/clds.Rda")

if (rs$regime_shift_name %in% clds$regime_shift) {
    gg <- rs_net(clds, rs$regime_shift_name) |> 
    plot_net()
    girafe(ggobj = gg)}

```


```{r}
load("assets/rs_analysis_text.Rda")

if (is.null(rs_analysis[[rs$regime_shift_name]]))  "This regime shift does not have a feedback analysis yet" else 
    rs_analysis[[rs$regime_shift_name]] |> writeLines()
```


## Citation

Acknowledge this review as:

> <small> `r str_c(rs$main_contributors, ", ",  rs$other_contributors, ". ", rs$regime_shift_name, "In: Regime Shift Database, www.regimeshifts.org. Last revised: ", rs$date)` </small>

## References

<small>

```{r results='asis'}
rs |> pull(references) |> 
    str_split(pattern = "\\(\\#\\d{1,}\\) ") |> 
    map(~str_c("- ", ., "\n")) |> 
    unlist() |> 
    str_remove(pattern = "- \\\n") |> 
    cat()
```

</small>
