---
output: 
  html_document:
    toc: false
    toc_depth: 3
    toc_float: true
bibliography: "assets/bibliography/rs2.bib"
params:
  j: 2
---


```{r include = FALSE}
library(tidyverse)
library(leaflet)

load('assets/cases_db.Rda')
cases <- dat
dat <- read_csv('assets/generic_types_RSDB_new.csv') |> 
    janitor::clean_names()

x <- dat$id[params$j]
rs <- dat |> filter(id== x)
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```


::::{style='display: flex;'}

:::{style='width: 50%;'}

# Indian summer monsoon


- **Main contributors**: Juan Carlos Rocha, Rolands Sadauskis

- **Other contributors**: Reinette (Oonsie) Biggs, Garry Peterson

- Last update: 2011-02-23

Globally there are six Monsoon systems: the Sub-Saharan Africa monsoon, the North American monsoon, the Indian summer monsoon also called South Asian monsoon (Southeast & Northeast monsoons), the East Asian monsoon, the Indo-Australian monsoon, and the European monsoon.   
  
The Indian Summer Monsoon has attracted most concern and is of importance to India's economy which is particularly reliant on the monsoon precipitation. It accounts for 80% of mean precipitation of India but it has decrease 4-5% in the last five decades (Bollasina et al. 2011). A regime shift from Indian summer monsoon with mean and regular precipitation within the season towards a weakened monsoon precipitation may be occurring in response to changes in vegetation driven by agricultural change. Changes in vegetation cover are leading to a decrease in the temperature difference between land and ocean that is essential for the creation of monsoon precipitation. Weaker monsoon reduce rains with strong consequences for agricultural productivity in one of the densely populated areas of the world.

 Several feedback mechanisms have been proposed to help and maintain a weakened monsoon precipitation. The primary and best understood is the moisture-advection feedback mechanism where decreasing release of latent heat from the lack of precipitation over land adds to the decrease in temperature difference between land and ocean. This drives weaker winds from ocean to land thus decreasing landward advection of moisture, which leads to further reductions of precipitation.



:::

:::{style='width: 5%;'}

:::

:::{style='width: 45%;'}

```{r carousel, out.width='500px', out.height='100px', results = 'asis'}



library(htmltools)
library(bsplus)

w <- 'https://www.juanrocha.se/'
bs_carousel(id = 'rs2', use_indicators = TRUE)|>
bs_append(
    content = bs_carousel_image(
        src = paste0(w, 'img/3972976144_b2912a1d8f_b.jpg'), alt = 'Indian summer monsoon'),
    caption = bs_carousel_caption(title = 'Indian summer monsoon'))|>
bs_append(
    content = bs_carousel_image(
        src = paste0(w, 'img/Monsoon_CLD.jpg'), alt = 'Indian summer monsoon'),
    caption = bs_carousel_caption(title = 'Indian summer monsoon'))

```

:::
::::

## Evidence

```{r map, out.width='100%', out.height='300px', results = 'markup'}



cs <- cases |> 
    filter(type == rs$regime_shift_name) 

m <- leaflet(cs) |> 
    addTiles(paste0('https://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png?=key=', keyring::key_get('carto-maps'))) |> 
    addCircleMarkers(
        lng = ~long, lat=~lat, radius = 2, 
        color = 'orange', 
        popup = ~popups) |> 
    setView(5,10, zoom = 2) 
m
```

**Fig 1 | Empirical evidence.** The databse currently documents `r nrow(cs)` cases of `r str_to_lower(rs$regime_shift_name)` around the world. Each dot in the map is coded from a scientific article documenting a place undergoing this regime shift. You can learn more from each case by hovering and clicking on the case of interest.

## Analysis {.tabset .tabset-pills}

### General information

#### Alternative regimes

The Indian Monsoon system is solar heating driven and starts in the southernmost point of the Indian Peninsula where it splits into two parts. One branch is moving along western part of India northwards. The other branch flows over the Bay of Bengal heading towards North-East India and Bengal, picking up more moisture from the Bay of Bengal. Later the winds arrive at the Eastern Himalayas with following heavy rainfall. After the arrival at the Eastern Himalayas, the winds turn West, covering northern and eastern India, Bangladesh, parts of Pakistan, and southern Nepal pouring rain all along its way.

 It is essential to recognize that there has already been a series of abrupt but opposite changes in monsoon intensity on centennial to millennial time scales, when the East Asian Monsoon rapidly strengthened while the Indian Summer Monsoon rapidly weakened [Bing et al 2006]. Though several studies have indicated that the East Asian summer monsoon has also become weaker after the end of the 1970s (Wang et. al 2001). Therefore, the end of 1970s may be viewed as an abrupt and significant change point in the inter-decadal variability of the East Asian summer monsoon [Ding et al. 2008]. Thus this change of both monsoon systems weakening at the same time might also point to a shift in the interconnected monsoon system. The decrease in precipitation in the Indian summer monsoon system has not been occurring at a linear rate. The annual precipitation levels have been varying from very low to high and there is increasing trend in the rainfall during pre-monsoon and post-monsoon months. Evidence for the fact that a regime shift has already occurred is weak and there is controversy over when or if a tipping point will be crossed [Lenton 2008]. For example IPCC projections do not show obvious threshold behaviour this century [IPCC 2007].

 **Indian summer monsoon with mean and regular precipitation within the season**

 In this regime precipitation reaches its annual mean amount and ensures that agriculture receives enough water to sustain the production of crops and other goods. The rapid change in increase of CO2 concentrations in atmosphere has supported the existing regime increasing the sea surface temperature thus increasing the precipitation (see Fig.1). Indian Summer monsoon circulation is characterized by monsoon rains arriving by the start of June as warm air converges and ascends in the low pressure over the continent, leading to clouds and heavy rainfall. Monsoon rains advance northward as summer progresses, enhanced by forced ascent as the flow reaches the Himalayan foothills. In winter, this thermally direct circulation reverses as the land surface cools relative to the oceans. Rains continue till the end of September when the season ends. (Rickenbach et al. 2009)

  **Indian summer monsoon with weak and irregular precipitation within the season**

 This regime is characterized by weak precipitation causing droughts with the rain arriving much later than expected in June. The increasing emissions of fossil fuel SO2 and black carbon result in large atmospheric concentrations of black carbon and other aerosols generating atmospheric Brown clouds. This induces strong negative trends in surface solar radiation, surface evaporation, and summer monsoon rainfall. [Ramanathan 2005]. As the monsoon season starts precipitation can be irregular - raining in June and then having weak or no precipitation in July and returning in August.



#### Drivers and causes of the regime shift

The main direct driver of changing monsoon rains is deforestation associated with the indirect driver of increased food production. As a consequence of the vegetation loss due to deforestation the surface albedo is increased. Therefore the amount of reflected solar radiation increases due to high albedo, thus decreasing the temperature difference between ocean and land. All this leads to change in the main monsoon circulation mechanism that is responsible for precipitation in the region.

 The increasing concentration of CO2 in atmosphere is a driver that is being speculated about whether it maintains the current regime or pushes the system towards a regime with weaker precipitation. Researchers are still arguing which case is most likely and it depends on the source to present the influence of this driver. IPCC report (2007) project that carbon release from anthropogenic sources will continue increasing during the coming decades. There are studies indicating that emissions of greenhouse gases that alter the heat budget of the system and therefore the land-sea temperature contrast, could increase the monsoon intensity and/or variability (Knopf et al 2008; Kripalani et al. 2007). Nevertheless Palmer et al. (1992) pointed out that enhanced convection (the transfer of heat by the actual movement of the warmed matter) associated with the warm SST anomalies suppresses the monsoon rainfall.

 After these discussions the influence of this driver on the regime shift is speculative. As the input of green house gases in atmosphere is global, therefore it also regards to CO2 increase in atmosphere as a driver.



#### Impacts on ecosystem services and human well-being

Due to irregular precipitation one of the most essential and recognized provisioning services that could be lost is freshwater as the groundwater levels are becoming deeper thus drying up the soil. Food crops and livestock are two other provisioning services that are directly affected by this regime shift and linked with the freshwater service. Loosing these services would mean poor harvests and food shortages from lack of cattle among the rural population, which constitutes two thirds of India's total population. (Knopf et al. 2008) Timber production would also be affected, as barren and dried soils couldn't support the growth of trees. Also the fire frequency would increase damaging timber productivity even more.

 Furthermore couple of regulating services would be lost. One such would be climate regulation as a weak summer monsoon can change climate variability (Gordon et al. 2008). Air quality regulation as a service is endangered because of increased aerosol/dust and Brown cloud concentrations caused from biomass burning and industrial pollution. Regulation of water and soil erosion would be lost as monsoon rains ensure that soils are moist and inhabited with flora and fauna enough not to lose the fertile topsoil due to wind or other type of erosions. Water regulation would also be in danger, as lack of precipitation would alter the water cycle changing the typical water distribution in it.

 Biodiversity would rapidly decrease in the case of weak precipitation in long term. In the case of this regime shift human well-being would decrease in various ways. For example monsoon rains ensure that the turbines of hydroelectric power plants continue to turn. Therefore with lack of precipitation energy supply would be inconvenient affecting almost all sectors, which also causes delay in productions or increase in costing of products. Food crisis from the loss of food crops ecosystem service can lead to rapid inflation on food prices. In turn this can lead to large number of people suffering from hunger, as any adverse effect on farming will affect the purchasing power of the people as well. The hunger and rise of poverty could result in large numbers of people emigrating from the region [Barnet & Adger, 2007]. Lack of freshwater also could aggravate the sanitation and health issues that already haven't been completely solved. Crime levels could potentially rise as the depression among society would increase and the necessity for food would drive the people to support their families in any circumstances [Barnet & Adger, 2007].



#### Management options

**Options for preventing regime shift** 

 Options for preventing a weakening of the Indian summer monsoon circulation system primary relate to the sustainable management of the local and regional vegetation cover. The area of complete deforestation should be decreased and cropland area planning has to be in place. That has to be done in order to avoid rapid changes in surface albedo in large area that can change the existing feedback mechanisms. Sustainable water management planning has to be practised in order to avoid significant losses in water due to irrigation. New irrigation practises has to be considered for example drip irrigation and low pressure pivots in order to get more yield with less use of water. It has also been suggested that a better weather forecasting system for India would help people to better adapt strategies in times of droughts and floods induced by the monsoon variability (Source: Nature News Blog, Aug-2011).

 In the case of the mean Indian summer monsoon pattern, Greenhouse gas concentrations also have to be managed. Greenhouse gases influence the system and increase precipitation that can lead to large floods. Furthermore due to the pollution caused by combustion from cars, factories etc., the amount of black carbon in atmosphere has been increasing forming brown clouds that negatively influence the monsoon rainfall. Sudden reduction in air pollution without a concomitant reduction in global Greenhouse gas also can accelerate the warming in South Asia. The atmospheric brown clouds have masked the surface warming due to greenhouse gases. Thus in this case managers have to be careful and need to monitor the Indian summer monsoon circulation and precipitation as emphasizing only one driver can lead to even bigger change. Therefore it is essential to understand local and regional actions that may influence the feedback mechanisms.

  **Options for restoration of desirable regimes**

 Technology transfer could be a good initiative from developed countries as they can provide more advanced technological solutions and funding to developing countries to help accelerate reduction of GHG emissions and irrigation management. Investing in sustainable irrigation tools and supporting the industrial production industry for example with filters that decrease the amount of pollutants entering the atmosphere, techniques that are energy efficient.

 Overall the shift in the Indian summer monsoon circulation is considered to be irreversible if the changes in vegetation cover due to food production continue to increase.



### In-depth analysis

```{r cld, error = FALSE, message = FALSE, out.width='80%'}

#| fig.cap = '**Fig 2| Causal diagram.** Hover over the variables to see their names, orange dots are drivers, blue dots are variables inside feedback mechanisms. Red arrows represent positive causal relationships while blue arrows represent negative ones. Key feedbacks are described below.'

library(ggiraph)
library(htmlwidgets)
source('tools.R')
load('assets/clds.Rda')

if (rs$regime_shift_name %in% clds$regime_shift) {
    gg <- rs_net(clds, rs$regime_shift_name) |> 
    plot_net()
    # create the html obj: sizing not working
    girafe(ggobj = gg) |> 
        girafe_options(
            opts_tooltip(opacity = 0.7), opts_zoom(min =0.5, max =2),
        sizingPolicy(defaultWidth='100px', defaultHeight='50px'),
        opts_hover(css = 'fill:red;stroke:orange;r:5pt;'))
    }


```

```{r}

load('assets/rs_analysis_text.Rda')

if (is.null(rs_analysis[[rs$regime_shift_name]]))  'This regime shift does not have a feedback analysis yet' else 
    rs_analysis[[rs$regime_shift_name]] |> writeLines()
```

### Categorical summary

::::{style='display: flex;'}

:::{style='width: 50%;'}

#### Impacts

**Ecosystem type**:'

- Moist savannas & woodlands
- Drylands & deserts 


**Key ecosystem processes**:

- Water cycling


**Biodiversity**:

- Biodiversity


**Provisioning services**:

- Freshwater
- Food crops
- Livestock
- Wild animal and plant products
- Timber
- Fuel and fiber crops
- Hydropower


**Regulating services**:

- Air quality regulation
- Climate regulation
- Water regulation


**Cultural services**:

- Aesthetic values


**Human well-being**:

- Food and nutrition or Health (e.g. toxins or disease) or Livelihoods and economic activity or Cultural or Aesthetic and recreational values or Social conflict


**Links to other regime shifts**:

- Forest to savanna
- Dryland degradation


:::

:::{style='width: 50%;'}

#### Drivers

**Key drivers**:

- Vegetation conversion and habitat fragmentation or Harvest and resource consumption or Soil erosion & land degradation


**Land use**:

- Small-scale subsistence crop cultivation or Large-scale commercial crop cultivation or Extensive livestock production (natural rangelands) or Timber production


#### Key attributes

**Spatial scale**:

- Sub-continental (e.g. southern Africa or Amazon basin)


**Time scale**:

- Years
- Decades


**Reversibility**:

- Unknown


**Evidence**:

- Models
- Paleo-observation
- Contemporary observations


**Confidence: existence of the regime shift**

- Speculative – Regime shift has been proposed, but little evidence as yet


**Confidence: mechanisms underlying the regime shift**

- Contested – Multiple proposed mechanisms, reasonable evidence both for and against different mechanisms


:::

::::

### References

::: {#refs}

:::
### References old [Delete]

<small>

- Bing H,Qinghua L,& Yetang H. 2006. Interconnections between the Asian monsoon, ENSO, and high northern latitude climate during the Holocene. Chinese Science Bulletin. 51(18),2169-2177 
- Cohen EJ, 2003. Human Population: The Next Half Century. Science 302(5648),1172 - 1175 
- Ding YH, Sun Y, & Wang ZY. 2008. Inter-decadal variation of summer precipitation in East China and its association with decreasing Asian summer monsoon: part I—observed evidences. International Journal of Climatology 28,1139–1161. 
- Douglas EM, Niyogi D,  Frolking J,  Yeluripati B, Pielke Sr.RA,  Niyogi N, Vörösmarty CJ, and Mohanty UC.  2006. Changes in moisture and energy fluxes due to agricultural land use and irrigation in the Indian Monsoon Belt, Geophysical Research Letters 33,5. 
- Fasullo J. 2005. Atmospheric hydrology of the anomalous 2002 Indian summer monsoon, Mon. Weather Rev. 133,2996–3014. 
- Fennessy et al., 1994. The simulated Indian monsoon: A GCM sensitivity study. J. Climate 7,33–43. 
- Gasse F, Arnold M, Fontes J C, et al. 1991. A 13000-year climate record from western Tibet. Nature 353,742-745. 
- Gordon LJ, Peterson GD,  Bennett EM. 2008. Agricultural modifications of hydrological flows create ecological surprises. Trends Ecol. Evol. 23,211–219. 
- IPCC. 2007. Climate Change 2007: The Physical Science Basis. Cambridge University Press, New York. 
- Knopf B,  Zickfeld K,  Flechsig M, Petoukhov V. 2008. Sensitivity of the Indian Monsoon to Human Activities. Advances In Atmospheric Sciences 25(6),932–945. 
- Kripalani RH, Kulkarni JH, Sabade A,and Chaudhari HS. 2007. South Asian Summer Monsoon Precipitation Variability: Coupled Climate Simulations and Projections Under IPCC AR4. Theoretical and Applied Climatology 90(3-4),133-159. 
- McGuffie K,  Henderson-Sellers A,  Zhang H,Durbidge TB,  Pitman AJ. 1995. Global climate sensitivity to tropical deforestation. Global and Planetary Change 10,97-128. 
- Playán E, Salvador R, Faci JM, Zapata N, Martínez-Cob A, and Sánchez I. 2005. Day and night wind drift and evaporation losses in sprinkler solid-sets and moving laterals. Agricultural Water Management 76(3),139-159. 
- Ramanathan V, Chung C, Kim D, Bettge T, Buja L, Kiehl JT, Washington WM, Fu Q, Sikka DR, and Wild M. 2005. Atmospheric brown clouds: Impacts on South Asian climate and hydrological cycle. Proc. Natl. Acad. Sci. 102,5326–5333. 
- Rickenbach TM, Ferreira RN, Rickenbach JF, and Wright EA. 2009. Monsoon in the Americas: Opportunities and Challenges. Geography Compass 3(5),1625–1640. 
- Rosenfeld D, Rudich Y, and Lahav R. 2001. Desert dust suppressing precipitation: A possible desertification feedback loop. Proc.Nat., Acad. Sci. 98(11),5975–5980 
- Satheesh SK, and Ramanathan V. 2000. Large differences in tropical aerosol forcing at the top of the atmosphere and Earth’s surface. Nature 405, 60–63 
- Sinha SK, and Swaminathan MS. 1991. Deforestation, climate change and sustainable nutrition security: A case study of India. Climatic Change 19(1-2),201-209. 
- Sirocko F, Sarnthein M, Erlenkeuser H, et al. 1993. Century-scale events in monsoonal climate over the past 24000 years. Nature 364,322-324. 
- Wang B, and Lau KM, 2001. Interannual variability of the Asian summer monsoon: Contrasts between the Indian and the western North Pacific–east Asian monsoons. J. Climate 14,4073–4090. 
- Zickfeld K, Knopf B, Petoukhov V, Schellnhuber HJ. 2005. Is the indian summer monsoon stable against global change? Geophys Res Lett 32.

</small>

## Citation
Acknowledge this review as:

<tiny> 
``` {style='color: gray;'}
Juan Carlos Rocha, Rolands Sadauskis, Reinette (Oonsie) Biggs, Garry Peterson. Indian summer monsoon. In: Regime Shift Database, www.regimeshifts.org. Last revised: 2011-02-23
```

</tiny>

BibTeX citation:

<tiny><tiny> 
``` {style='color: gray;'}
@misc{
  author = {Juan Carlos Rocha and Rolands SadauskisReinette (Oonsie) Biggs and Garry Peterson},
  title = {Indian summer monsoon},
  url = {www.regimeshifts.org},
  howpublished = {Regime Shifts Database},
  publisher = {Stockholm Resilience Centre},
  institution = {Stockholm University}
}
```

</tiny>