---
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography: "assets/bibliography/rs3.bib"
link-citations: TRUE
params:
  j: 3
---

```{r include = FALSE}
library(tidyverse)
library(leaflet)

load('assets/cases_db.Rda')
cases <- dat
dat <- read_csv('assets/generic_types_RSDB_new.csv') |> 
    janitor::clean_names()

x <- dat$id[params$j]
rs <- dat |> filter(id== x)
knitr::opts_chunk$set(echo = FALSE, results = 'asis')
```



```{r carousel, out.width='500px', out.height='100px', , results = 'asis'}
library(htmltools)
library(bsplus)

w <- "https://www.juanrocha.se/img/"
bs_carousel(id = "corals_transitions", use_indicators = TRUE) |>
    bs_append(
        content = bs_carousel_image(
            src = paste0(w, "Coral_Reef.jpg"), alt = "Coral dominated reef"),
        caption = bs_carousel_caption("Coral dominated reef") ) |>
    bs_append(
        content = bs_carousel_image(
            src = paste0(w, "nur03507.jpg"), alt = "Macroalgae"),
        caption = bs_carousel_caption("Macroalgae")) |> 
bs_append(
    content = bs_carousel_image(
        src = paste0(w, "Corallimorpharia_3.jpg"), alt = "Corallimorpharia"),
    caption = bs_carousel_caption("Corallimorpharia") ) |>
bs_append(
    content = bs_carousel_image(src = paste0(w, "Goniopora_djiboutiensis_(Soft_coral).jpg"), alt = "Soft corals"),
    caption = bs_carousel_caption("Soft corals") ) |>
bs_append(
    content = bs_carousel_image(
        src = paste0(w, "Red_Tube_Sponge.jpg"), alt ="Sponge"),
    caption = bs_carousel_caption("Sponge") ) |>
bs_append(
    content = bs_carousel_image(
        src = paste0(w, "Sea_urchins_DSC00593.JPG"), alt = "Urchin barren"),
    caption = bs_carousel_caption("Urchin barren") ) 
```




# Coral transitions

**Main contributors**: Juan Carlos Rocha

**Other contributors**: Reinette (Oonsie) Biggs, Garry Peterson, Albert Norström

Last update: 2011-02-25


## Summary

Regime shifts in coral reefs typically involve a change in species dominance from hard corals (3D structure) to algal dominance.
Less commonly documented shifts include shifts from hard corals to soft coral dominance, corallimorpharians, urchin barrens or sponge dominance.
All of these regime shifts result in loss of diversity and structural complexity, and are typically triggered by a combination overfishing, pollution, diseases and climate change.
Loss of biodiversity and coral bleaching make coral systems more vulnerable to such stressors.

::::: {style="display: flex;"}
::: {style="width: 50%;"}
#### Impacts

**Ecosystem type**:'

-   Marine & coastal

**Key ecosystem processes**:

-   Soil formation

**Biodiversity**:

-   Biodiversity

**Provisioning services**:

-   Fisheries
-   Wild animal and plant products

**Regulating services**:

-   Water purification
-   Regulation of soil erosion
-   Pest & disease regulation
-   Natural hazard regulation

**Cultural services**:

-   Recreation
-   Aesthetic values
-   Knowledge and educational values
-   Spiritual and religious

**Human well-being**:

-   Food and nutrition
-   Livelihoods and economic activity
-   Security of housing & infrastructure
-   Cultural
-   Aesthetic and recreational values

**Links to other regime shifts**:

-   Freshwater eutrophication
-   Forest to savanna
-   Fisheries collapse
:::

::: {style="width: 50%;"}
#### Drivers

**Key drivers**:

-   Harvest and resource consumption
-   External inputs (e.g. fertilizers
-   pest control
-   irrigation)
-   Adoption of new technology (e.g. new fishing nets)
-   Species introduction or removal
-   Disease
-   Soil erosion & land degradation
-   Environmental shocks (e.g. fire
-   floods
-   droughts)
-   Global climate change

**Land use**:

-   Fisheries
-   Conservation
-   Tourism

#### Key attributes

**Spatial scale**:

-   Local/landscape (e.g. lake
-   catchment
-   community)

**Time scale**:

-   Years

**Reversibility**:

-   Hysteretic (difficult to reverse)

**Evidence**:

-   Models
-   Paleo-observation
-   Contemporary observations
-   Experiments

**Confidence: existence of the regime shift**

-   Well established – Wide agreement in the literature that the RS exists

**Confidence: mechanisms underlying the regime shift**

-   Well established – Wide agreement on the underlying mechanism
:::
:::::

## Evidence

```{r map, out.width='100%', out.height='300px', results = 'markup'}

#| fig.cap = "Cases of coral transition documented in the database."

cs <- cases |> 
    filter(type == rs$regime_shift_name) 

m <- leaflet(cs) |> 
    addTiles('https://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png') |> 
    addCircleMarkers(
        lng = ~long, lat=~lat, radius = 2, 
        color = 'orange', 
        popup = ~popups) |> 
    setView(5,10, zoom = 2) 
m
```
**Fig 1 | Empirical evidence.** The databse currently documents `r nrow(cs)` cases of coral transitions around the world. Each dot in the map is coded from a scientific article documenting a place undergoing this regime shift. You can learn more from each case by hovering and clicking on the case of interest.

## Analysis {.tabset .tabset-pills}

### General information

#### Alternative regimes

Regime shifts in coral systems are usually associated with a change in the species dominance of this ecological community, and consequent changes in the ecosystem structure.
Coral reefs are marine ecosystems, three-dimensional shallow-water structures dominated by sclearactinean or hard corals [@bellwood2004].
The most well documented regime shifts entail shifts from hard coral to fleshy seaweed (macroalgae) dominance.
However, shifts from hard coral to corallimorpharians dominance, soft coral dominance, sponge dominance, and urchin barrens states have also been documented [@norström2009].

**Coral dominated reefs**

Corals reefs are marine communities considered one of the most biodiverse and economically important ecosystems on the planet [@hoegh-guldberg2007].
Coral reefs are old calcareous three-dimensional structures created by corals - coral polyps colonies in symbiosis with microscopic algae.
This regime implies high biodiversity and three dimensional habitat complexity that offer shelter to other species.

Coral reefs occupy less than 0.5% of ocean floor but sustain up to 25% of marine biodiversity, and produce at least 10% of fish consumed by humans [@moberg1999].
Due to their three dimensional structure, coral reefs host up to 60 000 species of plants and animals [@moberg1999]; and this rich diversity makes coral reefs a target for tourism.
In over 100 countries, coral reefs are major natural and economic resources [@Goreau1994].
For example, coral reefs are one of the major attractions in the Caribbean.
In that region tourism is a major source of foreign currency.
In some Caribbean countries tourism accounts for up to half of nation's gross domestic product [@hoegh-guldberg2007].

Coral reefs can also reduce coastal erosion by reducing the impact of ocean currents and waves.
Consequently, intact coral reefs protect coastal infrastructure and valuable sandy beaches.
Furthermore, increases in erosion can result in the degradation or loss of other coastal ecosystems such as mangroves and sea-grasses areas, which provide valuable ecosystem services such as fish nurseries [@moberg1999; @hoegh-guldberg2007].

**Algae dominated reefs**

Algae dominated benthos appear when algae overgrow corals.
Submarine algal emerge and reduce coral growth, eventually leading to the loss of the habitat complexity of coral reefs [@bellwood2004].
Algae dominated reefs have lower biodiversity, fish production, and provide less protection from coastal erosion.
There is also less primary production on algae dominated reefs [@moberg1999].

**Coralimorpharians, soft corals and sponges**.

Although less documented, other regimes can dominate former coral reefs when dramatically disturbed [@bruno2009; @norström2009].
Systems dominated by coralimorpharians, soft corals, and sponges are alternative regimes where coral patches can fall when strongly disturbed [@norström2009].
Corallimorpharians, soft corals and sponges are thought to reduce habitat complexity and hence biodiversity.
Thus, one may expect to see fisheries production and tourism services reduced.
However, because such systems are not well studied it is unclear to what extent the ecosystem services provided by coral reefs are reduced.

**Urchin barrens**

Urchins are herbivores that live on coral reefs.
In the absence of predatory fish, urchins can become the dominant grazer on reefs and produce urchin barrens.
Urchin barrens inhibit coral growth, because urchins erode calcium carbonate from living and dead corals eroding reef structure.
When urchins are present at densities high enough that their erosion exceeds coral growth, urban barrens can form [@norström2009].

Urchin barrens are the regime with most severe habitat complexity loss, which reduces habitat for other species.
There is low production of fish, and the urchin populations are vulnerable to disease.
However, urchin can be a valuable species when harvested, therefore urchin barrens can support an urchin harvest.

#### Drivers and causes of the regime shift

Coral regime shifts are driven by multiple drivers, including overfishing, pollution, disease, global warming and ocean acidification [@bellwood2004; @bruno2009; @mumby2007].
The hard coral regime is more desirable than the other five regimes for most people, because of the inherent biological, social and economical importance of coral reefs.

All regime shifts from hard coral dominated to any of the alternative regimes are influenced by the same set of drivers.
However, differences in the resulting regime are produced by contextual features or the ordering of events.
For example, coral reef systems in the Indian Ocean are more likely to shift towards coralimorpharians, soft corals or sponges since algal grazing herbivores are more persistent than they are in the Caribbean, where macro-algae dominance is more common.
The role of different drivers is relatively consistent across cases, but variation in their impact, timing and interaction likely produce different types of regime shifts in different regions.

#### Impacts on ecosystem services and human well-being

A regime shift from a coral reef to another regime results in a decline in multiple ecosystem services including coastal erosion regulation, fisheries, tourism, water cleansing, and calcium fixation [@moberg1999].
Coral reefs occupy less than 0.5% of ocean floor but sustain up to 25% of its biodiversity and produce at least 10% of fish consumed by humans [@moberg1999].
Alternative regimes do not support biodiversity or fish to the same extent.
Constanza et al.[@costanza1997] estimated the value of coral reefs services at up to US\$ 6 075 per hectare per year based on their contribution to disturbance regulation, waste treatment, biological control, refuge habitat, food production, raw materials, recreation and cultural values.
Many of these services are lost or substantially reduced when regime shifts away from coral reefs occur.

Coral reef regime shifts can result in the collapse of coral fisheries that can produce unemployment for fishermen, and reduce the value of the fishery, as well as reduce food production.
Recreational services (primarily based around diving and snorkelling) are diminished when regime shifts occur, causing losses estimated at up to AUS\$ 682 million in the Great Barrier Reef and US\$ 8.9 billion for the Caribbean, in addition to 350 000 jobs related in the Caribbean [@moberg1999].
Coral reefs support cultural and spiritual values such as religious rituals, cultural traditions and institutional frameworks for cooperative fishing, especially in small scale fishing communities [@moberg1999].

#### Management options

**Options for enhancing resilience**

> *''The persistence of hard coral dominated reefscapes beyond 2050 will be heavily reliant on 2 things, the ability of corals to increase their upper thermal bleaching limits by \~0.1°C per decade, and management that produce local conditions that constrain excessive algal biomass proliferation during inter-disturbance intervals''* — [@wooldridge2005]

Temperature and acidification are global drivers that are difficult for local mangers to influence, yet the management of the local level drivers can make a difference [@hoegh-guldberg2007].
Due to the multi-causal nature of coral regime shifts, scholars emphasize the necessity of managing coral reefs using an ecosystem approach.
It requires taking into account the interaction between land and sea, as well as the scale and origin of the stressors when making decisions at different scales of governance [@moberg1999].
Management programs which successfully address the improvement of water quality, reduction of sediments, nutrients, toxins, pathogens and fishing pressure will increase the likelihood of corals to recover to shocks like bleaching events [@wooldridge2005; @hoegh-guldberg2007; @houk2010].

Herbivore populations are a key driver that can be actively managed because reduced grazing increases the vulnerability to regime shifts [@mumby2007].
For example, herbivorous fish like parrotfish can be protected.
Another option is to transform markets to incorporate a body of incentives to prevent the depletion of species in critical functional groups [@bellwood2004].
The abundance of sea urchins should also be carefully managed because urchin dominance can produce negative effects on coral recruitment [@norström2009].

The spatial resilience of coral reefs is a regional driver that is possible to manage by managing connectivity, metapopulation dynamics, and to take into consideration the spatial distribution of coral reefs [@moberg1999; @nyström2001].
This is because large-scale regional shifts are typically preceded by smaller-scale localized shifts.
Therefore, monitoring the occurrence and spatial distribution of smaller-scale regime shifts may help to anticipate, and potentially avert, large-scale catastrophic shifts [@norström2009].

To manage spatial resilience, Bellwood *et al.* [-@bellwood2004] recommend increasing the rate of establishment and size of no-take areas, including 'cool-spots' of biodiversity.
The reason is that areas with low species richness may be more vulnerable, as they may have lost functional groups, or may have low functional redundancy.
Hence, minor changes in such ecosystems may trigger regime shifts locally, and erode spatial resilience regionally.
International agreements are badly needed to keep oceans condition below the 480ppm and 2ºC thresholds for coral reefs [@hoegh-guldberg2007].

Reducing pressure from global scale drivers, requires coordinated global action, but would substantially increase the resilience of coral reefs.

### In depth analysis

```{r cld, error = FALSE, message = FALSE, out.width='80%'}
#| fig.cap = "**Fig 2| Causal diagram.** Hover over the variables to see their names, orange dots are drivers, blue dots are variables inside feedback mechanisms. Red arrows represent positive causal relationships while blue arrows represent negative ones. Key feedbacks are described below."
#| 
source('tools.R')
load('assets/clds.Rda')

if (rs$regime_shift_name %in% clds$regime_shift) {
    gg <- rs_net(clds, rs$regime_shift_name) |> 
    plot_net()
    girafe(ggobj = gg)}
```

```{r}
## use this to extract text, then comment out and format properly
# load('assets/rs_analysis_text.Rda')
# 
# if (is.null(rs_analysis[[rs$regime_shift_name]]))  'This regime shift does not have a feedback analysis yet' else 
#     rs_analysis[[rs$regime_shift_name]] |> writeLines()

```

####  Feedback mechanisms

The feedbacks maintaining each regime can be defined by the competitive relationships among corals, macroalgae, coralimorpharians, soft corals, sponges and urchin barrens *(local, proposed)*.
Each regime reinforce itself by occupying space and consuming nutrients that then become inaccessible to other species groups.

##### Coral dominated reefs

-   *Symbiosis with zooxanthellae feedback (local, well-established)*: The coral regime is reinforced by its symbiosis with zooxanthellae.
    Microscopic algae called zooxanthellae live inside the polyps.
    Corals receive food from the algae, which have the ability to photosynthesize.
    The algae also facilitate skeletal growth and provide corals with their color.
    In return, corals offer the algae nutrients (nitrogen, phosphorous and carbon dioxide) and protection from predators.
    When coral abundance is reduced by external disturbances, competitors may establish and restrict the ability of corals to regrowth.

##### Algae dominated reefs

-   *Unpalatability feedback (local, well-established):* If macroalgae reach certain size, it becomes unpalatable for herbivores [@scheffer2008; @hughes2010], creating a reinforcing feedback that favors further macroalgae recruitment.

##### Coralimorpharians, soft corals and sponge

-   *Competition feedback (local, well-established):* Coralimorpharians, soft corals and sponges are set of functional groups of species that usually compete with corals for space and resources.
    After strong disturbance events such as hurricanes, bleaching events, El Niño or La Niña events, low tides, disease outbreaks, oil pollution or euthropication that reduce coral populations, coral reefs are prone to be overgrown by its competitors.

##### Urchin barrens

-   *Bioerosion feedback (local, well-established):* The formation of urchin barrens usually happens in environments with low biodiversity and absence of predators.
    Urchin are herbivores in coral reefs, but when they form barrens, they may predate corals as well.
    The urchin barren regime is established when the bioerosion rate is higher than the reef accretion rate [@norström2009].

#### Drivers

##### Shift from Coral dominated reefs to alternative regimes

**Important shocks** (eg droughts, floods) that contribute to the regime shift include:

-   *Thermal anomalies (regional, well-established):* Thermal anomalies produce coral bleaching, a phenomena seen when corals release its zooxanthellae due to thermal stress, leading to the loss of coral color.
    If temperatures exceed summer maxima by 1º to 2ºC for over 3 weeks, then bleaching will occur, with more severe bleaching as thermal anomalies intensify and lengthen [@hoegh-guldberg2007].
    In the Great Barrier Reef, Australia, the maximum sea surface temperature (SST) over any 3 day period during the bleaching season (summer) predicted the presence/absence of bleaching with 73.2% accuracy [@berkelmans2004].
    Massive events of coral bleaching have been recorded more recently [@berkelmans2004].
    According to model predictions, coral bleaching events are expected to become increasingly frequent and severe in the coming decades due to global warming [@wooldridge2005; @hoegh-guldberg2007].
    Corals may survive and recover from bleaching after mild thermal stress, but typically show reduced growth, calcification, and fecundity and may experience greater incidences of coral disease [@bruno2007; @hoegh-guldberg2007].

-   *Hurricanes (regional, contested):* Hurricanes are shock events that can destroy coral structure, reducing coral populations.
    These shocks can provide opportunities for algal populations to dominate over coral reefs, especially when these shocks occur in synergy with other drivers.
    However, hurricanes might also have a beneficial role when passing several hundred kilometers from coral reefs as they potentially cool down water from thermal anomalies [@eakin2010].

-   *Coral disease outbreaks (local to regional, contested):* Coral disease outbreaks may also act in synergy with other drivers by weakening coral’s resilience to other shocks.
    The same works the other way around.
    For example, warmer water temperature and nutrients inputs have been correlated with increase of disease outbreaks (e.g. white band syndrome) [@bruno2007; @houk2010].
    However, in cases where disease probability is density dependent, outbreaks are less likely to happen.
    It depends on the disease dynamics, transmission patterns and contextual factors.
    This is why it is considered a contested driver.

-   *Low tides (local, well-established):* Low tides have been reported to induce mass coral mortality [@norström2009].
    They expose the coral to higher solar radiation, luminosity and temperature.

-   *Oil spills (local, controversial):* Oil spills reduce luminosity inhibiting photosynthesis, change the chemical composition of water, and may induce coral mortality directly by the effect of toxins or by affecting larvae survival and dispersal.

The main **external direct drivers** that contribute to the shift include:

-   *Fishing (local, well-established):* Overfishing reduces herbivory, leading to increasing macroalgal abundance due to reduced grazing [@mumby2007].
    Fishing can also reduce the diversity of herbivores.
    When coral cover is reduced there is increased colonization by algae, which in turn inhibits coral recruitment – i.e., a positive feedback exists [@mumby2007; @norström2009].
    When the abundance of fish is reduced, the ability to react to algal growth peaks is reduced as well, and mats of algae may establish and cover portions of a coral reef.

-   *Atmospheric* $CO_2$ *(Global, well-established):* ocean acidification reduces the concentration of carbonate-ions in the oceans, which is a fundamental compound required for corals to grow.
    With less carbonate accretion corals cannot fix their skeletons, hence the coral reef structure will be reduced.
    Many coral are expected to be unable to cope with $CO_2$ concentrations above 480ppm and temperature increase above +2ºC [@hoegh-guldberg2007].

-   *Global warming (Global, well-established):* Global warming is expected to increase sea surface temperature, furthermore by increasing temperature it also increases $CO_2$ concentration in water and therefore ocean acidification.

The main **external indirect drivers** that contribute to the shift include:

-   *Deforestation (regional, proposed):* Deforestation increases the leakage of nutrients and runoff of sediments from soils to water bodies.
    Sedimentation and nutrients inputs are in turn a source of stress for coral reefs and by fertilizing algal populations stimulate algal growth while decreasing coral growth rates [@hoegh-guldberg2007].

-   *Urbanization (local, proposed):* Similar to deforestation, coastal development increase the amount of sewage on waterbodies, leading to more sedimentation and nutrient inputs, which stimulates algae at the expense of corals [@hoegh-guldberg2007].

-   *Food demand (global, well-established):* Food demand is a driver that stimulates fishing that in turn is a direct driver for coral degradation [@salvat1992].
    Market connectivity, trade facilities, government policies and development of technologies allow local fishermen to increase fishing effort [@deutsch2011].

**Slow internal system changes** that contribute to the regime shift include:

-   *Herbivory (local, well-established):* Herbivores (mainly fish) control the growth of algae through grazing, scraping, and bioeroding [@nyström2001].
    However, once algae grow beyond a certain size they become unpalatable to herbivores [@scheffer2008].
    If the abundance of fish is reduced, the ability to react to algal growth is reduced as well.
    Consequently, reduction in herbivore diversity decreases the ability of herbivores to regulate temporary increases algal growth due to disturbance [@nyström2001].
    Diseases that attack herbivore species can initiate a positive feedback that can lock the system into an alternative regime dominated by macroalgae [@nyström2001].

-   *Connectivity loss (regional, well-established):* For all types of regime shifts described, there is a common pattern in regard to spatial resilience.
    Spatial connectivity is needed to allow larvae interchange, which increase genetic variability.
    Well-connected reefs are more resilient to disturbances [@nyström2001].
    However, when connectivity is broken, corals rely on self-seeding and are more vulnerable to depletion of local stocks, bleaching events and other disturbances [@elmhirst2009; @hughes2010].

#### Key thresholds

-   If temperatures exceed summer maxima by 1º to 2ºC for over 3 weeks, then bleaching will occur, with more sever bleaching as thermal anomalies intensify and lengthen [@hoegh-guldberg2007]

-   Coral reefs are expected to decline if carbon dioxide exceeds 480ppm and temperature increase +2ºC [@hoegh-guldberg2007]

### References

::: {#refs}
:::

<br> <br>

## Citation

Acknowledge this review as:

<tiny>

``` {style="color: gray;"}
Juan Carlos Rocha, Reinette (Oonsie) Biggs, Garry Peterson, Albert Norström. Coral transitions. In: Regime Shift Database, www.regimeshifts.org. Last revised: 2011-02-25
```

</tiny>

BibTeX citation:

<tiny>

``` {style="color: gray;"}
@misc{rocha2011corals,
  author = {Rocha, Juan and Biggs, Reinette Oonsie, and Peterson, Garry and Norström, Albert},
  title = {Coral transitions},
  url = {www.regimeshifts.org},
  year = {2011},
  howpublished = {Regime Shifts Database},
  publisher = {Stockholm Resilience Centre},
  institution = {Stockholm University}
}
```

</tiny>
